Type-Safe Modular Parsing

### 2017-09-04 SLE_17/sections/Inheritance.tex
shortening.

\begin{lstlisting}
class Var(x: String) extends Expr {
  def print = x
}
\end{lstlisting}
Here one may quickly define a new parser \lstinline|pVar| in \lstinline|AParser| for variables, and parse new expressions
with \lstinline{"pExpr ||| pVar"}.
Unfortunately, even \inlinecode{"1 + x"} cannot be parsed, which is obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to \lstinline|pExpr| for parsing sub-expressions, whereas the newly added \inlinecode{pVar} is not observed, unless we replace all the occurrences of \lstinline|pExpr| with \lstinline{"pExpr ||| pVar"}.
Yet modifying existing code breaks semantic modularity.

@@

\lstinputlisting[linerange=20-26]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_APPROACH

@@

\vspace{-.05in}
\paragraph{Independent Extensibility}
A nice feature of Scala is its support for the linearized-style multiple
inheritance on traits~\cite{odersky2004overview}. This can be very helpful when
composing several languages, and to achieve independent
extensibility~\cite{odersky2005independently}. Suppose now we want to compose the parsers
for expressions from pre-defined languages \lstinline|LanguageA| and \lstinline|LanguageB| using alternative.
The new parser can be built by inheriting both parsers at the same time:
\begin{lstlisting}
trait LanguageA {...}
trait LanguageB {...}
trait LanguageC extends LanguageA with LanguageB {
  override def pExpr = super[LanguageA].pExpr ||| super[LanguageB].pExpr
}
\end{lstlisting}

>>>>

\vspace{-4pt}
\begin{lstlisting}
class Var(x: String) extends Expr {
  def print = x
}
\end{lstlisting}
\vspace{-4pt}

Here one may quickly define a new parser \lstinline|pVar| in \lstinline|AParser| for variables, and parse new expressions
with \lstinline{"pExpr ||| pVar"}.
Unfortunately, even \inlinecode{"1 + x"} cannot be parsed, which is obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to \lstinline|pExpr| for parsing sub-expressions, whereas the newly added \inlinecode{pVar} is not observed, unless we replace all the occurrences of \lstinline|pExpr| with \lstinline{"pExpr ||| pVar"}.
Yet modifying existing code breaks semantic modularity.

@@

\vspace{-4pt}
\lstinputlisting[linerange=20-26]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_APPROACH
\vspace{-4pt}

@@

\vspace{-.05in}
\paragraph{Independent Extensibility}
A nice feature of Scala is its support for the linearized-style multiple
inheritance on traits~\cite{odersky2004overview}. This can be very helpful when
composing several languages, and to achieve independent
extensibility~\cite{odersky2005independently}. Suppose now we want to compose the parsers
for expressions from pre-defined languages \lstinline|LanguageA| and \lstinline|LanguageB| using alternative.
The new parser can be built by inheriting both parsers at the same time:
\vspace{-4pt}
\begin{lstlisting}
trait LanguageA {...}
trait LanguageB {...}
trait LanguageC extends LanguageA with LanguageB {
  override def pExpr = super[LanguageA].pExpr ||| super[LanguageB].pExpr
}
\end{lstlisting}\vspace{-4pt}

### 2017-09-04 SLE_17/sections/ObjectAlgebras.tex
shortening.

\lstinputlisting[linerange=23-23]{code/src/papercode/Sec4OA/Code1.scala}% APPLY:linerange=FVARSEXPR

@@

\lstinputlisting[linerange=5-8]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
Here \inlinecode{Alg} is called an \textit{Object Algebra interface}, parameterized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

@@

\lstinputlisting[linerange=12-16]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can also define an evaluation operation as a new trait that extends \inlinecode{Alg[Int]}. Hence adding new operations is modular. We omit that code
due to space reasons.

@@

\lstinputlisting[linerange=27-29]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALGEXT
Now pretty-printing on the new language can be realized without modifying existing code:

\lstinputlisting[linerange=33-35]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
Thus existing code was reused and was not modified!

@@

\lstinputlisting[linerange=42-43]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP

@@

\lstinputlisting[linerange=8-17]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser looks fine, but it is not extensible. For example, we have demonstrated in Section~\ref{sec:inheritance} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]}, which is a \emph{supertype} of the old type \inlinecode{Alg[E] =>} \inlinecode{Parser[E]}, since
the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This violates
overriding in Scala.

@@

\lstinputlisting[linerange=34-39]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER

@@

\lstinputlisting[linerange=46-49]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

>>>>

\vspace{-4pt}
\lstinputlisting[linerange=23-23]{code/src/papercode/Sec4OA/Code1.scala}% APPLY:linerange=FVARSEXPR
\vspace{-4pt}

@@

\vspace{-4pt}
\lstinputlisting[linerange=5-8]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
\vspace{-4pt}
Here \inlinecode{Alg} is called an \textit{Object Algebra interface}, parameterized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

@@

\vspace{-4pt}
\lstinputlisting[linerange=12-16]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
\vspace{-4pt}
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can also define an evaluation operation as a new trait that extends \inlinecode{Alg[Int]}. Hence adding new operations is modular. We omit that code
due to space reasons.

@@

\vspace{-4pt}
\lstinputlisting[linerange=27-29]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALGEXT
\vspace{-4pt}
Now pretty-printing on the new language can be realized without modifying existing code:

\vspace{-4pt}
\lstinputlisting[linerange=33-35]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
\vspace{-4pt}
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
Thus existing code was reused and was not modified!

@@

\vspace{-4pt}
\lstinputlisting[linerange=42-43]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP
\vspace{-4pt}

@@

\vspace{-4pt}
\lstinputlisting[linerange=8-17]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
\vspace{-4pt}
Such a parser looks fine, but it is not extensible. For example, we have demonstrated in Section~\ref{sec:inheritance} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]}, which is a \emph{supertype} of the old type \inlinecode{Alg[E] =>} \inlinecode{Parser[E]}, since
the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This violates
overriding in Scala.

@@

\vspace{-4pt}
\lstinputlisting[linerange=34-39]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER
\vspace{-4pt}

@@

\vspace{-4pt}
\lstinputlisting[linerange=46-49]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT
\vspace{-4pt}

### 2017-09-04 SLE_17/sections/PackratParsing.tex
Spell check. Bibtex.
shortening.

\section{Packrat Parsing for Modularity}\label{sec:packrat}

@@

\paragraph{Longest Match Composition} Another problematic issue
in parser combinator libraries is the need for manually prioritizing/ordering
alternatives in a grammar.
Consider the grammar:
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <int> `+' <expr>
\end{grammar}

@@

\paragraph{Backtracking} The need for backtracking can also be problematic
in a modular setting. Consider a grammar that includes \inlinecode{"import..from"}, 
and is extended with an \inlinecode{"import..as"} case:

\setlength{\grammarindent}{5em}
\begin{grammar}
<stmt> ::= `import' <ident> `from' <ident>
    \alt ...
    \alt `import' <ident> `as' <ident>
\end{grammar}
Since the two cases share a common prefix, when the former fails, we must backtrack to the beginning.
For example, the choice combinator in Parsec only tries the second alternative if the first fails
without any token consumption. We have to use \inlinecode{try} for explicit backtracking.

\begin{lstlisting}[language=PlainCode]
oldParser = parseImpFrom <|> ...
newParser = try parseImpFrom <|> ... <|> parseImpAs
\end{lstlisting}

Similarly, this violates a modular setting because it also requires a global view of the full grammar.
Hence the worst case where all alternatives may share common prefixes with future cases
should always be antecipated. Therefore we need to backtrack for all the branches. To avoid failures in the future, we have to add \inlinecode{try} everywhere. However this results in the worst-case exponential time
complexity.

@@

\lstinputlisting[linerange=9-15]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_OO_AST

@@

\lstinputlisting[linerange=19-26]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_SIMPLE_EXPR

@@

\lstinputlisting[linerange=6-7]{code/src/papercode/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_RUNPARSER

>>>>

\vspace{-2pt}
\section{Packrat Parsing for Modularity}\label{sec:packrat}

@@

\paragraph{Longest Match Composition} Another problematic issue
in parser combinator libraries is the need for manually prioritizing/ordering
alternatives in a grammar.
Consider the grammar:
\vspace{-2pt}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <int> `+' <expr>
\end{grammar}
\vspace{-2pt}

@@

\paragraph{Backtracking} The need for backtracking can also be problematic
in a modular setting. Consider a grammar with \inlinecode{"import..from"}, 
and is extended with an \inlinecode{"import..as"} case:

\vspace{-2pt}
\setlength{\grammarindent}{5em}
\begin{grammar}
<stmt> ::= `import' <ident> `from' <ident>
    \alt ...
    \alt `import' <ident> `as' <ident>
\end{grammar}
\vspace{-2pt}
Since the two cases share a common prefix, when the former fails, we must backtrack to the beginning.
For example, the choice combinator in Parsec only tries the second alternative if the first fails
without any token consumption. We have to use \inlinecode{try} for explicit backtracking.

\vspace{-4pt}
\begin{lstlisting}[language=PlainCode]
oldParser = parseImpFrom <|> ...
newParser = try parseImpFrom <|> ... <|> parseImpAs
\end{lstlisting}
\vspace{-4pt}

Similarly, this violates a modular setting because it also requires a global view of the full grammar.
Hence the worst case where all alternatives may share common prefixes with future cases
should always be anticipated. Therefore we need to backtrack for all the branches. To avoid failures in the future, we have to add \inlinecode{try} everywhere. However this results in the worst-case exponential time
complexity.

@@

\vspace{-4pt}
\lstinputlisting[linerange=9-15]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_OO_AST
\vspace{-4pt}

@@

\vspace{-4pt}
\lstinputlisting[linerange=19-26]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_SIMPLE_EXPR
\vspace{-4pt}

@@

\vspace{-4pt}
\lstinputlisting[linerange=6-7]{code/src/papercode/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_RUNPARSER
\vspace{-10pt}

### 2017-09-04 SLE_17/sections/RelatedWork.tex
shortening.

\section{Related Work}\label{sec:relatedwork}

@@

Besides, type-specific languages~\cite{omar14} do support
separate compilation and type safety. The modularity of parsing is incorporated in its type
system, which allows users to introduce new syntax for parsing. Whereas in our approach, parsing 
strategy is an orthogonal issue since we rely on an external parsing library.
Our approach does not rely on an elaborated semantics extension, but simply uses a pattern
that is applicable in mainstream OO languages, where the type system helps ensure
type safety.

@@

\paragraph{Extensible Parsing Algorithms}
\textit{Parse table composition}~\cite{bravenboer2008parse, schwerdfeger2010}
is an approach where grammars are compiled to
modular parse tables. Those parse tables are expressed as DFAs
or NFAs, and later they can be composed by an algorithm, to provide
separate compilation for parsing. The generation of parse tables can
be quite expensive in terms of performance. The approach
is quite different from ours, since it uses parse
 tables, whereas we use parser combinators.
Our approach supports both
\emph{separate compilation} as well as \emph{modular
  type-checking}, and is applied in common OO languages. Moreover, the extensibility of parsing is further
available at language composition.

@@

\paragraph{Parser Combinators} Parser combinators have become more and more
popular since~\cite{burge1975,Wadler1985}. Many parsing libraries produce recursive descent
parsers by introducing functional monadic
parser combinators~\cite{nott237}. Parsec~\cite{Leijen2001} is
perhaps the most popular parser combinator library in this line.
It is widely used in Haskell (with various ``clones'' in other languages)
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination,
high cost for backtracking and longest match composition issues,
as we discussed in Section~\ref{subsec:challenges}. Those limitations make Parsec
(and similar parsing techniques) inadequate for modular parsing.

@@

\paragraph{Extensibility} Various design patterns~\cite{gamma1995design} in multiple
languages, have been proposed over the years to address extensibility
problems, such as the Expression Problem~\cite{wadler1998expression}.
The famous ``Datatypes \`a
la Carte'' (DTC)~\cite{swierstra2008data} approach represents modular ASTs using co-products
of every two functors. Several variants of DTC have been later proposed~\cite{Bahr2011,Bahr2014,Oliveira2015}.
All of that work essentially covers how to traverse and
consume extensible ASTs. However they do not
address the problem of \emph{modularly
parsing extensible ASTs}. Only in Bahr's~\cite{Bahr2011} work \emph{unfolds} is briefly mentioned,
yet it does not cover parsing.

@@

It is worth mentioning that Scala case classes~\cite{emir2007matching} provide a near
solution to the Expression Problem. Case classes can be
modularly added, but they do not enforce
exhaustiveness of pattern matching for extensible operations. In other
words run-time pattern matching errors can happen when writing extensible code with case classes. So that full static type-safety is not ensured. Nevertheless case classes are a pragmatic approach, which is
widely used in practice. Therefore, modular parsing techniques may be
of value for extensible code using case classes.  The approach we
presented in Section~\ref{sec:inheritance} can readily be adapted to case classes:
all that the users need to do is to use case classes, instead of
standard OO classes in their code.

>>>>

\vspace{-3pt}
\section{Related Work}\label{sec:relatedwork}

@@

Besides, type-specific languages~\cite{omar14} do support
separate compilation and type safety. The modularity of parsing is incorporated in its type
system, which allows users to introduce new syntax for parsing. Whereas in our approach, parsing 
strategy is based on an external parsing library.
Our approach does not rely on an elaborated semantics extension, but simply uses a pattern
that is applicable in mainstream OO languages, where the type system ensures
type safety.

@@

\vspace{-6pt}
\paragraph{Extensible Parsing Algorithms}
\textit{Parse table composition}~\cite{bravenboer2008parse, schwerdfeger2010}
is an approach where grammars are compiled to
modular parse tables. Those parse tables are expressed as DFAs
or NFAs, and later they can be composed by an algorithm, to provide
separate compilation for parsing. The generation of parse tables can
be quite expensive in terms of performance. The approach
is quite different from ours, since it uses parse
 tables, whereas we use parser combinators.
Our approach supports both
\emph{separate compilation} as well as \emph{modular
  type-checking}, and is commonly applicable OO languages. Moreover, the extensibility of parsing is further
available at language composition.

@@

\vspace{-3pt}
\paragraph{Parser Combinators} Parser combinators have become more and more
popular since~\cite{burge1975,Wadler1985}. Many parsing libraries produce recursive descent
parsers by introducing functional monadic
parser combinators~\cite{nott237}. Parsec~\cite{Leijen2001} is
perhaps the most popular parser combinator library in this line.
It is widely used in Haskell (with various ``clones'' in other languages)
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination,
high cost for backtracking and longest match composition issues,
as we discussed in Section~\ref{subsec:challenges}. Those limitations make Parsec
(and similar parsing techniques) inadequate for modular parsing.

@@

\vspace{-3pt}
\paragraph{Extensibility} Various design patterns~\cite{gamma1995design} in multiple
languages, have been proposed over the years to address extensibility
problems, such as the Expression Problem~\cite{wadler1998expression}.
The famous ``Datatypes \`a
la Carte'' (DTC)~\cite{swierstra2008data} approach represents modular ASTs using co-products
of every two functors. Several variants of DTC have been later proposed~\cite{Bahr2011,Bahr2014,Oliveira2015}.
All of that work essentially covers how to traverse and
consume extensible ASTs. However they do not
address the problem of \emph{modularly
parsing extensible ASTs}. Only in Bahr's~\cite{Bahr2011} work \emph{unfolds} is briefly mentioned,
yet it does not cover parsing.

@@

It is worth mentioning that Scala case classes~\cite{emir2007matching} provide a near
solution to the Expression Problem. Case classes can be
modularly added, but they do not enforce
exhaustiveness of pattern matching for extensible operations. In other
words run-time pattern matching errors can happen when writing extensible code with case classes. So that full static type-safety is not ensured. Nevertheless case classes are a pragmatic approach, which is
widely used in practice. Therefore, modular parsing techniques may be
of value for extensible code using case classes.  The approach we
presented in Section~\ref{sec:inheritance} can readily be adapted to case classes:
all that the users need to do is to use case classes, instead of
standard OO classes in their code.
\vspace{-5pt}
